How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-07-20 2.12-taxonomy.Rmd
Add files via upload
# After copy edits

There is only one valid scientific name for a species. This is expressed in a binomial: two names, first the genus (with an initial capital letter) and then the species name (lower case). This name is italicised by convention (or in some cases underlined). Note that no other taxonomic names (e.g. Family, Order, Class, Phylum) are written with italics (although they all get capital first letters), only the genus and species are italicised. The single name is a hard rule as taxonomy can't survive in any other way. The most recently published version is the one to use, but happily with many vertebrates [mammals](https://www.mammaldiversity.org/), [birds](https://avibase.bsc-eoc.org/avibase.jsp?lang=EN), [fish](https://www.fishbase.se/search.php), [reptiles](http://www.reptile-database.org/) and [amphibians](http://research.amnh.org/vz/herpetology/amphibia/) there are well curated databases to which you can refer for the latest taxonomic treatments. Also see [WoRMS](http://www.marinespecies.org/) for marine species. My suggestion is that you defer to these (with citation) and you won't go far wrong. If you do this early on in your thesis, remember that you may need to update it later.

@@

When my former student, Mohlamatsane Mokhatla, started his PhD in 2013 he worked on three species, and over time these changed so that at various points in his thesis. You can see the combinations he had to contend with in Table \@ref(tab:taxonomy).

@@

The taxonomy of two species changed such that he had to constantly revise the names in his thesis. Luckily, he could simply conduct a "replace all", but it did become quite confusing. There's a chance that they've changed again, so I've provided hyperlinks for you to check out what the current name is. The lesson is that if you want to avoid this, work on *Xenopus laevis* or conduct your PhD in as short a time as possible!

In taxonomic papers, and certain journals with a taxonomic focus, there is an insistence that you cite the "taxonomic authority" after the species name when used for the first time. This is essentially a citation in which you acknowledge the original description that accompanies the name. Let's take Mohlamatsane's examples again:

@@

It might seem that there's a mistake above, because Tschudi, 1838 isn't in brackets. This is actually deliberate. Tschudi, 1838 isn't in brackets because the genus name hasn't changed since the description by Tschudi [1838](tschudi1838classification) placed *S. capensis* in *Sclerophrys* in 1838. But it is clear in the table above that the genus name did change to *Amietophrynus*, and it was recognised under a different species name (*A. rangeri*). Unlikely as it may seem, the original description of the species was by Tschudi in 1838, but this was long forgotten (or ignored because the type specimen was so poorly preserved), and the species was regularly referred to as *Bufo rangeri*. When the genus *Bufo* was broken up into lots of newly named genera by Frost et al [-@frost2006amphibian], Tschudi's description was once again disregarded [@ohler2016identity]. However, in 2016 new work on the type specimen revealed that it was the same as *Amietophrynus rangeri,* but because Tschudi's description was older, the taxonomy deferred to the rule of precedence. If you want to read more about it, you can look at all the changes on the AMNH amphibian database.

@@

The generic name of a Latin binomial is often abbreviated to the first letter (e.g. *X. laevis*). This should only be done once it is well established in your writing what the full name is. Or, in other words, you should refer to the species in full on the first mention, and then switch to the abbreviation. Because sentences shouldn't start with an abbreviation, you might need to relax this from time to time. You also need to treat the abstract as a stand-alone document in this respect. You might find that you have two species, which belong to different genera that share the same first letter. In this case, most authors use a different abbreviation: for example *Ap. rangeri* and *At. angolensis*. This is exasperated when both species names are the same. Keen readers would have noticed that Mohlamatsane doesn't have this problem any more!

@@

 - Viruses – [International Code of Virus Classification and Nomenclature](https://talk.ictvonline.org/information/w/ictv-information/383/ictv-code)
 - Bacteria & Archaea [International Code of Nomenclature of Prokaryotes](https://www.the-icsp.org/bacterial-code)
 - Algae, Fungi & Plants [International Code of Nomenclature for algae, fungi, and plants](https://www.iapt-taxon.org/nomen/main.php)
 - Animals [International Code of Zoological Nomenclature](https://www.iczn.org/the-code/the-international-code-of-zoological-nomenclature/)

@@

Plant taxonomy is considered the taxonomic mother discipline as the work of Linneaus in 1753 was the first botanical work that consistently used Linneus' binomial (two names: Genus species) nomeclatural system [@linnaeus1753species] which pre-dated the more inclusive (and better known) landmark publication in 1758 [@linnaeus1758systema]. Until the Botanical Code was revised in 2012, all new botanical diagnoses had to be written in Latin. In addition to the Latin binomial name (Genus and species), plants (and some other taxa) often require further descriptors or **ranks** that denote subspecies, varieties, subvareties, etc. I have included a (non-exhaustive) list of these in Table \@ref(tab:plantrank), together with some examples.

Unlike animal authorities, for plants if the combination of Genus and species has changed, both the authority for the original genus placement, and the author responsible for the new combination are given after the species name (with the former in brackets). In addition, for plants no year is given, and the name of the authority is usually abbreviated. Because Linneaus named so many plant species is his landmark publications, instead of writing out the full taxonomic authority after the species names, botanists abbreviate this to "L.". Like Linneaus, other prolific plant taxonomists can have their names abbreviated, but plant taxonomists became so numerous that there needed to be some standardisation of names and their abbreviations. A database of these **Standard Forms** can be found at [the International Plant Names Index database](https://www.kew.org/data/authors.html). Note that Standard Forms in plant taxonomy can include initials.

Table: (\#tab:plantrank) **Plants often need more information than the simple Latin biomial.** This table provides the required abbreviations that you will need to add in order to properly describe the rank of your plant taxa.

| Rank | Abbreviation	| Example |
|-----|-----|-----|
| Subfamily | subf. | Poaceae subf. Pooideae |
| Subgenus | subg. | _Metrosideros_ subg. _Metrosideros_ (Myrtaceae) |
| Section | sect. | _Senecio_ sect. _Senecio_ (Asteraceae) |
| Subsection | subsect. | _Stachys_ subsect. _Swainsonianeae_ (Labiatae) |
| Series | ser. | _Mimosa_ ser. _Paucifoliolatae_ |
| Subseries | subser. | _Palicourea_ subser. _subcymosae_ |
| Subspecies | subsp. | _Trichophorum cespitosum_ subsp. _germanicum_ |
| Variety | var. | _Closterium acerosum_ var. _elongatum_ |
| Subvariety | subvar. | _Hieracium vulgatum_ subvar. _oblongatum_ |
| Form | f. | _Prunus cerasifera_ f. _atropurpurea_ |

Algae sometimes exist in association with other taxa (such as lichens which are an association between an algae or cyanobacteria and one or more fungi. For these associations there is a specific code for their nomenclature: International Code of Phytosociological Nomenclature [@theurillat2021international].

@@

@@

What name to use - This is tricky. My suggestion would be to take the most commonly used common name, as that's likely to be the one that most people will recognise. However, it's not always that simple and because there's no "correct" common name, you have some leeway to choose the one you prefer.

@@

There has been a recent movement to formulate common names into genera (or even families) such that the specific common name takes on a specific epithet of the generic or familial common name. In a world of stable taxonomy (such as that enjoyed by ornithologists), I can understand that this is possible. But imagine the requirements in common name changes for Mohlamatsane in his thesis over only a few years. Surely the useful thing about common names is that they are used commonly by normal people who don't worry about taxonomy? If we insist on common names constantly changing, don't we risk alienating the public?

@@

### View 1.

@@

### View 2.

>>>>

There is only one valid scientific name for a species. This is expressed in a binomial: two names, first the genus (with an initial capital letter) and then the species name (lower case). This name is italicised by convention (or, in some cases, underlined). Note that no other taxonomic names (e.g. Family, Order, Class, Phylum) are written with italics (although they all get capital first letters), only the genus and species are italicised. The single name is a hard rule as taxonomy can't survive in any other way. The most recently published version is the one to use, but happily with many vertebrates ([mammals](https://www.mammaldiversity.org/), [birds](https://avibase.bsc-eoc.org/avibase.jsp?lang=EN), [fish](https://www.fishbase.se/search.php), [reptiles](http://www.reptile-database.org/) and [amphibians](http://research.amnh.org/vz/herpetology/amphibia/)) there are well curated databases to which you can refer for the latest taxonomic treatments. Also, see [WoRMS](http://www.marinespecies.org/) for marine species. My suggestion is that you defer to these (with citation) and you won't go far wrong. If you do this early on in your thesis, remember that you may need to update it later.

@@

When my former student started his PhD in 2013, he worked on three species, and over time these changed so that at various points in his thesis they had a variety of names. You can see the combinations he had to contend with in Table \@ref(tab:taxonomy).

@@

The taxonomy of two species changed such that he had to constantly revise the names in his thesis. Luckily, he could simply conduct a 'replace all', but it did become quite confusing. There's a chance that they've changed again, so I've provided hyperlinks for you to check out what the current name is. The lesson is that if you want to avoid this, work on *Xenopus laevis* or conduct your PhD in as short a time as possible!

In taxonomic papers, and certain journals with a taxonomic focus, there is an insistence that you cite the 'taxonomic authority' after the species name when used for the first time. This is essentially a citation in which you acknowledge the original description that accompanies the name. Let's take the above examples again:

@@

It might seem that there's a mistake above, because Tschudi, 1838 isn't in brackets. This is actually deliberate. Tschudi, 1838 isn't in brackets because the genus name hasn't changed since the description by Tschudi [1838](tschudi1838classification) placed *S. capensis* in *Sclerophrys* in 1838. But it is clear in the table above that the genus name did change to *Amietophrynus*, and it was recognised under a different species name (*A. rangeri*). Unlikely as it may seem, the original description of the species was by Tschudi in 1838, but this was long forgotten (or ignored because the type specimen was so poorly preserved), and the species was regularly referred to as *Bufo rangeri*. When the genus *Bufo* was broken up into lots of newly named genera by Frost et al [-@frost2006amphibian], Tschudi's description was once again disregarded [@ohler2016identity]. However, in 2016 new work on the type specimen revealed that it was the same as *Amietophrynus rangeri,* but because Tschudi's description was older, the taxonomy deferred to the rule of precedence. If you want to read more about it, you can look at all the changes on the amphibian taxonomic database.

@@

The generic name of a Latin binomial is often abbreviated to the first letter (e.g. *X. laevis*). This should only be done once it is well established in your writing what the full name is. Or, in other words, you should refer to the species in full on the first mention, and then switch to the abbreviation. Because sentences shouldn't start with an abbreviation, you might need to relax this from time to time. You also need to treat the abstract as a stand-alone document in this respect. You might find that you have two species, which belong to different genera that share the same first letter. In this case, most authors use a different abbreviation: for example, *Ap. rangeri* and *At. angolensis*. This is exasperated when both species' names are the same.

@@

 - Viruses – [International Code of Virus Classification and Nomenclature](https://talk.ictvonline.org/information/w/ictv-information/383/ictv-code)
 - Bacteria & Archaea – [International Code of Nomenclature of Prokaryotes](https://www.the-icsp.org/bacterial-code)
 - Algae, Fungi & Plants – [International Code of Nomenclature for algae, fungi, and plants](https://www.iapt-taxon.org/nomen/main.php)
 - Animals – [International Code of Zoological Nomenclature](https://www.iczn.org/the-code/the-international-code-of-zoological-nomenclature/)

@@

Plant taxonomy is considered the taxonomic mother discipline as the work of Linneaus in 1753 was the first botanical work that consistently used Linneus' binomial (two names: Genus species) nomenclatural system [@linnaeus1753species] which pre-dated the more inclusive (and better known) landmark publication in 1758 [@linnaeus1758systema]. Until the Botanical Code was revised in 2012, all new botanical diagnoses had to be written in Latin. In addition to the Latin binomial name (Genus and species), plants (and some other taxa) often require further descriptors or **ranks** that denote subspecies, varieties, subvarieties, etc. I have included a (non-exhaustive) list of these in Table \@ref(tab:plantrank), together with some examples.

Unlike animal authorities, for plants, if the combination of Genus and species has changed, both the authority for the original genus placement, and the author responsible for the new combination are given after the species name (with the former in brackets see Table Table \@ref(tab:plantrank)). In addition, for plants no year is given, and the name of the authority is usually abbreviated. Because Linneaus named so many plant species in his landmark publications, instead of writing out the full taxonomic authority after the species names, botanists abbreviate this to "L.". Like Linneaus, other prolific plant taxonomists can have their names abbreviated, but plant taxonomists became so numerous that there needed to be some standardisation of names and their abbreviations. A database of these **Standard Forms** can be found at [the International Plant Names Index database](https://www.kew.org/data/authors.html). Note that Standard Forms in plant taxonomy can include initials.

Table: (\#tab:plantrank) **Plants often need more information than the simple Latin binomial.** This table provides the required abbreviations that you will need to add in order to properly describe the rank of your plant taxa. Note that the taxonomic authority (last column) can include two names with the first in parentheses if the second changed the placement of this taxon.

| Rank | Abbreviation	| Example | Authority |
|-----|-----|-----|-----|
| Subfamily | subf. | Poaceae subf. Pooideae | Benth.  |
| Subgenus | subg. | _Metrosideros_ subg. _Metrosideros_ (Myrtaceae) |   |
| Section | sect. | _Senecio_ sect. _Senecio_ (Asteraceae) |   |
| Subsection | subsect. | _Stachys_ subsect. _Swainsonianeae_ (Labiatae) |   |
| Series | ser. | _Mimosa_ ser. _Paucifoliolatae_ | Benth.  |
| Subseries | subser. | _Palicourea_ subser. _subcymosae_ | (Müll.Arg.) C.M.Taylor  |
| Subspecies | subsp. | _Trichophorum cespitosum_ subsp. _germanicum_ | (Palla) Hegi |
| Variety | var. | _Closterium acerosum_ var. _elongatum_ | Brebisson |
| Subvariety | subvar. | _Hieracium vulgatum_ subvar. _oblongatum_ | Sudre |
| Form | f. | _Prunus cerasifera_ f. _atropurpurea_ | Diffel |

Algae sometimes exist in association with other taxa (such as lichens which are an association between an algae or cyanobacteria and one or more fungi. For these associations, there is a specific code for their nomenclature: International Code of Phytosociological Nomenclature [@theurillat2021international].

@@

## Other times to use italics

In general, italics are used when names are derived from Latin or Greek. A number of subdisciplines have international codes for how the names should be written: Anatomy, Histology, Embryology and Neuroanatomy. There is often confusion about these correct terms, but luckily four major publications are here to help you. These are maintained by the international groups who specialise in standardising the nomenclature, and they are republished from time to time (so look out for updates). I provide directions to these citations should your research require them:

- Terminologia Anatomica. [@terminology1998terminologia]
- Terminologia Histologica. [@terminology2008terminologiaa]
- Terminologia Embryologica. [@terminology2013terminologia]
- Terminologia Neuroanatomica. [@terminology2017terminologia]

@@

What name to use - This is tricky. My suggestion would be to take the most commonly used common name, as that's likely to be the one that most people will recognise. However, it's not always that simple and because there's no 'correct' common name, you have some leeway to choose the one you prefer.

@@

There has been a recent movement to formulate common names into genera (or even families) such that the specific common name takes on a specific epithet of the generic or familial common name. In a world of stable taxonomy (such as that enjoyed by ornithologists), I can understand that this is possible. But imagine the requirements in common name changes for frogs in Table \@ref(tab:taxonomy) over only a few years. Surely the useful thing about common names is that they are used commonly by normal people who don't worry about taxonomy? If we insist on common names constantly changing, don't we risk alienating the public?

@@

### View 1

@@

### View 2

### 2021-07-20 2.13-style.Rmd
Add files via upload
# After copy edits

> “The goal of good writing is straightforward: to make your reader's job as easy as possible” 
> ---  Kevin W. Plaxco [-@plaxco2010art].

This great quote from a paper by Plaxco [-@plaxco2010art], who cuts to the heart of why it’s essential to write well. As an editor, I once experienced an author who thought the opposite, and for me this was the epiphany. The submitted manuscript was an impenetrable mess. I could tell that there was good work in there, but as an editor I felt that the information had been made so obscure by the authors that my readers were unlikely to get much out of it. The author was a colleague, and so I decided to phone him and chat about the need for much greater clarity in his manuscript. Experience with email has taught me that they can often be taken in the wrong way; usually in the worst possible way.

The response I got surprised me. The author recognised that his text was dense and was unapologetic. “Let the reader work to make sense of my data,” he said. My epiphany came not because this was a totally alien concept, but because during my PhD days this was exactly how I had thought scientific writing should be done. I had spent so many hours slogging through dense and dreary papers by well-respected figures. The reward came when I finally understood what it was that they were trying to communicate. This felt so great that I believed the puzzle they had set me was what I should set my own readers. Happily, my advisor didn’t feel the same way, but for years I continued to believe that a paper should be a decent puzzle for readers to crack. This might seem a crazy concept, but it’s not an uncommon problem. For a great take on this in Chemistry, see this article by Murray [-@murray2011skillful].

The preponderance of inpenetrable guff has led to a general perception that "academics stink at writing" - as Steven Pinker put it [@pinker2014why]. But, I think that Pinker [-@pinker2014why] is outdated. Certainly, as Plaxco [-@plaxco2010art] has it, there are a lot of ways that academics can improve their writing, for each other. But the belief that all academic writing is going to be accessible to the general public is far fetched. This is because there are a lot of terms when writing journal articles that are there for precision and brevity. These terms will exclude a larger audience, especially when they get so dense that looking them up is simply too arduous. What Plaxco [-@plaxco2010art] urges us to do is to make the rest of the text as accessible and easy as possible.

@@

There are still authors out there who attempt to set puzzles for their readers, but they aren’t in the mainstream any more. Instead, biological sciences has some inspiring writers, and many more are taught how to inspire future generations towards communicating great science. That’s not to say that there are no puzzles left. Much of what we do requires great puzzle solving abilities. However, let’s keep the puzzles away from communicating with our audience, make their job as reader easy or even pleasurable, and they’ll keep coming back for more.

## Don’t bamboozle

It’s easy to use jargon. The whole point of jargon is to convey a (usually complex) idea in a short amount of space. Using a word (or two) instead of using several sentences clearly has some advantages. However, there is such a thing as too much jargon. Simply put, it’s unnecessary to use jargon when you can use plain English in the same amount of space. My old advisor at Liverpool University, the late, great Brian Moss, shared the following example of too much jargon by Bump et al [@bump2009large], when plain English would have been much shorter. Do yourself a favor and read the abstract of Bump et al [-@bump2009large]. The fact that it gave Moss a chance to write about moose wasn’t lost on anyone (see Figure \@ref(fig:moss))!

(ref:moss) **Bamboozled Moss takes on the moose.** Brian was so unhappy after reading Bump et al (2009) the above that he felt compelled to write a letter to the British Ecological Society Bulletin.

@@

Whatever your writing style, you need to be conciious of your ultimate aim in writing: to communicate with your reader. Making your text as easy to read as possible [as @plaxco2010art put it] isn't easy, but it will be rewarding. The reward you will get is in willing readers to the message that you want to convey. This is something that will stand you in good sted no matter what career you have after your PhD.

>>>>

> “The goal of good writing is straightforward: to make your reader's job as easy as possible.” 
>
> Kevin W. Plaxco [-@plaxco2010art]

This great quote from a paper by Plaxco [-@plaxco2010art], who cuts to the heart of why it’s essential to write well. As an editor, I once experienced an author who thought the opposite, and for me, this was the epiphany. The submitted manuscript was an impenetrable mess. I could tell that there was good work in there, but as an editor, I felt that the information had been made so obscure by the authors that my readers were unlikely to get much out of it. The author was a colleague, and so I decided to phone him and chat about the need for much greater clarity in his manuscript. Experience with email has taught me that they can often be taken in the wrong way; usually in the worst possible way.

The response I got surprised me. The author recognised that his text was dense and was unapologetic. “Let the reader work to make sense of my data,” he said. My epiphany came not because this was a totally alien concept, but because during my PhD days, this was exactly how I had thought scientific writing should be done. I had spent so many hours slogging through dense and dreary papers by well-respected figures. The reward came when I finally understood what it was that they were trying to communicate. This felt so great that I believed the puzzle they had set me was what I should set my own readers. Happily, my advisor didn’t feel the same way, but for years I continued to believe that a paper should be a decent puzzle for readers to crack. This might seem a crazy concept, but it’s not an uncommon problem. For a great take on this in Chemistry, see this article by Murray [-@murray2011skillful].

The preponderance of impenetrable guff has led to a general perception that "academics stink at writing" - as Steven Pinker put it [@pinker2014why]. But, I think that Pinker [-@pinker2014why] is outdated. Certainly, as Plaxco [-@plaxco2010art] has it, there are a lot of ways that academics can improve their writing, for each other. But the belief that all academic writing is going to be accessible to the general public is far fetched. This is because there are a lot of terms when writing journal articles that are there for precision and brevity. These terms will exclude a larger audience, especially when they get so dense that looking them up is simply too arduous. What Plaxco [-@plaxco2010art] urges us to do is to make the rest of the text as accessible and easy as possible.

@@

There are still authors out there who attempt to set puzzles for their readers, but they aren’t in the mainstream any more. Instead, biological sciences have some inspiring writers, and many more are taught how to inspire future generations towards communicating great science. That’s not to say that there are no puzzles left. Much of what we do requires great puzzle-solving abilities. However, let’s keep the puzzles away from communicating with our audience, make their job as reader easy or even pleasurable, and they’ll keep coming back for more.

## Develop your writing style

During the course of your PhD you should improve your writing style toward proficient communication with your target audience. This will involve improving both your technical writing and accessibility writing skills. Although the bottom left portion of Figure \@ref(fig:develop-writing-style) is labelled “Poor communication” we might more productively think of it as being the area of the ‘first draft’. You have to start somewhere. If your current style is too ‘hard’ and technical (top left), then you will need to increase your accessibility. If on the other hand you are used to a more popular writing style (bottom right), you will need to increase your technical ability. See Dunleavy [-@dunleavy2003authoring] for more insight on achieving your requisite writing style.

(ref:develop-writing-style) **Develop your scientific writing style.** Pitching your writing style will take some practice to reach the desired pitch. Wherever your starting point in this technicality and accessibility square, your aim is to move toward the top right corner. Adapted from Dunleavy (2003: Fig. 5.1).

```{r develop-writing-style, echo=FALSE, out.width = '95%', fig.cap="(ref:develop-writing-style)"}
knitr::include_graphics('figures/develop-writing-style.png')
```

### Don’t bamboozle {#bamboozle}

It’s easy to use technical jargon (top left portion of Figure \@ref(fig:develop-writing-style)). The whole point of jargon is to convey an (usually complex) idea in a short amount of space. Using a word (or two) instead of using several sentences clearly has some advantages. However, there is such a thing as too much jargon. Simply put, it’s unnecessary to use jargon when you can use plain English in the same amount of space. My old advisor at Liverpool University, the late, great Brian Moss, shared the following example of too much jargon by Bump et al [@bump2009large], when plain English would have been much shorter. Do yourself a favor and read the abstract of Bump et al. [-@bump2009large]. The fact that it gave Moss a chance to write about moose wasn’t lost on anyone (see Figure \@ref(fig:moss))!

(ref:moss) **Bamboozled Moss takes on the moose.** Brian was so unhappy after reading Bump et al. (2009) that he felt compelled to write a letter to the British Ecological Society Bulletin.

@@

Whatever your writing style, you need to be conscious of your ultimate aim in writing: to communicate with your reader. Making your text as easy to read as possible [as @plaxco2010art put it] isn't easy, but it will be rewarding. The reward you will get is having willing readers understand the message that you want to convey. This is something that will stand you in good stead no matter what career you have after your PhD.

### 2021-07-20 2.14-voice.Rmd
Add files via upload
# After copy edits

# When is it possible to retain your own voice when writing? {#voice}

Should one use “I” or “we” when writing a scientific paper? Although there was a tradition not to use first-person pronouns when writing scientific articles, this has fallen by the wayside in recent years, because the use of I or we makes writing simpler for writers and easier to understand for readers. Use “I” if you are the sole author (or in your thesis) and “we” for two or more authors.

You will regularly see “I” or “we” being used in the last paragraph of the introduction and the first paragraph of the discussion. It’s also quite common to use this in the methods section. The reason why these sections have “I” or “we” is quite obvious, as it allows you to place your own voice on your actions and decisions. After all, it is your aim and hypothesis for your study, so you should own it. Using first-person pronouns also allows you to be more concise, and there are some great examples given by Kurt Vonnegut [-@vonnegut1980how] where you can see what happens when you don’t use them.

The first use of “I” or “we” at the end of the introduction is powerful. Up to that point, the reader has read four or five paragraphs of logical arguments outlining the background information that makes up the reasoning for your study. Using the first-person pronoun now makes the reader sit up as they have you communicating directly with them what you aimed to do. But that power is lost with repetition. So if you were to continuously use it throughout the methodology, the reader may get tired of reading a string of statements stating that you did this, then you did that, and so on. The power of using “I” or “we” is to use it sparingly to boost clarity.

## Does writing “I” or “we” help people to understand who you are?

@@

Having said that you should write about subjects that you feel passionately about, this does not mean that you should write with unsubstantiated passion. For example, it makes no sense to write: "The threats to this spectacular species surpass any mechanism known to humankind" ...unless you have substantive citations or results to back this up. Yes, it's passionate, but that should be balanced by the need to keep what you write factual.

>>>>

# Retaining your own voice when writing {#voice}

Should one use first-person pronouns ('I' or 'we') when writing a scientific paper? Although there was a tradition not to use first-person pronouns when writing scientific articles, this has fallen by the wayside in recent years, because the use of I or we makes writing simpler for writers and easier to understand for readers. Use 'I' if you are the sole author (or in your thesis) and 'we' for two or more authors.

You will regularly see 'I' or 'we' being used in the last paragraph of the introduction and the first paragraph of the discussion. It’s also quite common to use this in the methods section. The reason why these sections have 'I' or 'we' is quite obvious, as it allows you to place your own voice on your actions and decisions. After all, it is your aim and hypothesis for your study, so you should own it. Using first-person pronouns also allows you to be more concise, and there are some great examples given by Kurt Vonnegut [-@vonnegut1980how] where you can see what happens when you don’t use them.

The first use of 'I' or 'we' at the end of the introduction is powerful. Up to that point, the reader has read four or five paragraphs of logical arguments outlining the background information that makes up the reasoning for your study. Using the first-person pronoun now makes the reader sit up as they have you communicating directly with them what you aimed to do. But that power is lost with repetition. So if you were to continuously use it throughout the methodology, the reader may get tired of reading a string of statements stating that you did this, then you did that, and so on. The power of using 'I' or 'we' is to use it sparingly to boost clarity.

## Does writing 'I' or 'we' help people to understand who you are?

@@

Having said that you should write about subjects that you feel passionately about, this does not mean that you should write with unsubstantiated passion. For example, it makes no sense to write: "'"The threats to this spectacular species surpass any mechanism known to humankind" ...unless you have substantive citations or results to back this up. Yes, it's passionate, but that should be balanced by the need to keep what you write factual.

### 2021-07-20 2.15-concise.Rmd
Add files via upload
# After copy edits

It is likely, that like me, you chose to do science at school instead of useful subjects such as English and/or other languages. It might have also occurred to you that it was a really bad idea not to have done more earnest work in languages now that you are expected to write like a professional. Certainly, if you did pay more attention in languages in school, you may not be suffering with the rest of us now. Personally, I never received tuition in the grammatical construction of English (you may know this already). Thus I'm left in the dark most of the time about exactly what is wrong. Instead, I simply try to rework a sentence into what I know is correct (and I often don't achieve that). The lesson then is that should you come across an opportunity to learn more about writing, do take it.

Use writing professional emails (to your advisor, co-authors or collaborators) as an opportunity to write concise messages that communicate efficiently and precisely the aim of your message. As a PhD student, you should consider yourself a professional and these communications to colleagues will be received with more import if you can write concisely. A personal plea from me would be to ask that you don't start your email by enquiring after the health of your recipients. The fashion to to this has turned into a surge in recent years (the pandemic notwithstanding), and the reality is that this are simply superfluous and gratuitous words. You certainly don't expect them to write back starting with a paragraph about their health. Save this kind of correspondence for your friends and relatives, who may well appreciate you enquiring about their health. When your advisor, co-authors or collaborators become friends, you can move to a less formal style, but I'd argue that you always want to keep your email communication concise.

Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing. Maybe write a letter (or long email) instead of making a call sometimes. It'll also help you to [read critically](critical.html). If you are a keen reader, then you could do worse than reading some Kurt Vonnegut [-@vonnegut1980how].

@@

It's important to avoid distracting your reader (see [Part II](part2.html)). Your aim is to be thorough, which will mean including all relevant information, but don't allow your writing to sideline them into taking the wrong direction. You need to stick to your narrative like a highway. The highway analogy is useful as although you will see signs to other places that you can point out (to the reader), you shouldn't turn off the highway. The best of highways will also be free of traffic jams (think long, complex sentences here). and toll free ([no paywall](paywall.html)).

@@

If the highway is 'your message' then getting to the point is the big sign that states the destination.

Some of the best papers I've read manage to encapsulate the whole point in the first sentence, or better still in the title. Our formula has you getting to the main question in the last paragraph of the introduction, but you should have already 'got to the point' in the first paragraph - i.e. the point that is the bigger picture.

@@

You can find a guide to writing your methods in [Part III](part3.html). This rule is about not allowing the methods to creep out of this section and into the results, or even the discussion. Similarly with the Results. There are some specific times when this is permitted (such as a post-hoc test), but generally you shouldn't expect to do this.

@@

Copy and pasting is very easy, and a good way of suddenly producing vast quantities of text (Figure \@ref(fig:plagiarism)). But repeated text is quickly recognised by the reader and appears very boring and cumbersome. This is likely to happen in the Methods and Results sections. If you find yourself deciding that you can simply cut and paste this paragraph while changing the variable names and the numbers, then you are wrong. Don't ever do it.

@@

People also have a tendency to build the [abstract](abstract.html) by copying and pasting text from the main sections. Don't do it. The reader will quickly see the repetitions and become bored. Similarly, the [conclusion/summary](discussion.html) section is also often copied from lines above. If it's not worth writing again, then it wasn't worth writing before. Make it fresh, and keep it interesting!

@@

Interesting. I'd say that you need some practice and a critical eye to spot these kinds of errors. Having your advisor help or someone who has edited a lot of text. Scott's got some great examples [@hotaling2020simple].

@@

For some reason, most students avoid this at all cost. There appears to be the idea that saying "I" or "we" isn't correct for scientific writing. In fact, it can be the easiest way to avoid the passive voice. There are other reasons why [using the first person](voice.html) is helpful.

@@

This is really getting down to the nitty gritty. It's hard to do this yourself. It's much easier for someone else to show you. Typically I only do this level of word-smithing for abstracts or when the imposed word limit means that you really need to remove excess words. However, Scott is correct that if you can learn to do this yourself, it will improve your writing.

@@

I spend a lot of my time reading and commenting on the writing of my students. Every comment is made to improve the document. I do get mad when I find that this help is ignored without an indication why. My suggestions or comments are simply deleted. Certain students become very hard work when they don't even explain why they refuse to change something. Reading the work of these students quickly becomes a chore. You should have noted by now that it's all about being flexible, and being stubborn about anything you've written isn't going to work.

>>>>

It is likely, that like me, you chose to do science at school instead of useful subjects such as English and/or other languages. It might have also occurred to you that it was a really bad idea not to have done more earnest work in languages now that you are expected to write like a professional. Certainly, if you did pay more attention in languages in school, you may not be suffering with the rest of us now. Personally, I never received tuition in the grammatical construction of English (you may know this already). Thus I'm left in the dark most of the time about exactly what is wrong. Instead, I simply try to rework a sentence into what I know is correct (and I often don't achieve that). The lesson is that should you come across an opportunity to learn more about writing, do take it. Otherwise, try writing concisely in all that you do.

Use writing professional emails (to your advisor, co-authors or collaborators) as an opportunity to write concise messages that communicate efficiently and precisely the aim of your message. As a PhD student, you should consider yourself a professional and these communications to colleagues will be received with more import if you can write concisely. A personal plea from me would be to ask that you don't start your email by enquiring after the health of your recipients. The fashion to do this has turned into a surge in recent years (the pandemic notwithstanding), and the reality is that these are simply superfluous and gratuitous words. You certainly don't expect them to write back starting with a paragraph about their health. Save this kind of correspondence for your friends and relatives, who may well appreciate you enquiring about their health. When your advisor, co-authors or collaborators become friends, you can move to a less formal style, but I'd argue that you always want to keep your email communication concise.

Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing. Maybe write a letter (or long email) instead of making a call. It'll also help you to [read critically](critical.html). If you are a keen reader, then you could do worse than reading some Kurt Vonnegut [-@vonnegut1980how].

@@

It's important to avoid distracting your reader (see [Part II](part2.html)). Your aim is to be thorough, which will mean including all relevant information, but don't allow your writing to sideline them into taking the wrong direction. You need to stick to your narrative like a highway. The highway analogy is useful as although on a highway you will see signs to other places that you can point out to the reader, you shouldn't turn off the highway to get to your destination. The best of highways will also be free of traffic jams (think long, complex sentences here), and toll free ([no paywall](paywall.html)).

@@

If the highway is 'your message' then getting to the point is the big sign that states that you've arrived at your destination.

Some of the best papers I've read manage to encapsulate the whole point in the first sentence, or better still, in the title. Our formula has you getting to the main question in the last paragraph of the introduction, but you should have already 'got to the point' in the first paragraph - i.e. the point that is the bigger picture.

@@

You can find a guide to writing your methods in [Part III](#methods). This rule is about not allowing the methods to creep out of this section and into the results, or even the discussion. Similarly with the Results. There are some specific times when this is permitted (such as a post-hoc test), but generally, you shouldn't expect to do this.

@@

Copy and pasting are very easy, and a good way of suddenly producing vast quantities of text (Figure \@ref(fig:plagiarism)). But the repeated text is quickly recognised by the reader and appears very boring and cumbersome. This is likely to happen in the Methods and Results sections. If you find yourself deciding that you can simply cut and paste this paragraph while changing the variable names and the numbers, then you are wrong. Don't ever do it.

@@

People also have a tendency to build the [abstract](abstract.html) by copying and pasting text from the main sections. Don't do it. The reader will quickly see the repetitions and become bored. Similarly, the [conclusion/summary](discussion.html) section is also often copied from the lines above. If it's not worth writing again, then it wasn't worth writing before. Make it fresh, and keep it interesting!

@@

This rule relates to keeping your text concise, specifically at the start of a sentence. You may be tempted to reiterate your point, but this is not always needed. I'd say that you need some practice and a critical eye to spot these kinds of errors. Having your advisor help or someone who has edited a lot of text. Scott's got some great examples [@hotaling2020simple].

@@

For some reason, most students avoid this at all cost. There appears to be the idea that saying 'I' or 'we' isn't correct for scientific writing. In fact, it can be the easiest way to avoid the passive voice. There are other reasons why [using the first person](#voice) is helpful.

@@

This is really getting down to the nitty-gritty. It's hard to do this yourself. It's much easier for someone else to show you. Typically I only do this level of word-smithing for abstracts or when the imposed word limit means that you really need to remove excess words. However, Scott is correct that if you can learn to do this yourself, it will improve your writing.

@@

I spend a lot of my time reading and commenting on the writing of my students. Every comment is made to improve the document. I do get mad when I find that this help is ignored without an indication why. Occasionally, my suggestions or comments are simply deleted. Certain students become very hard work when they don't explain why they refuse to change something. Reading the work of these students quickly becomes a chore. You should have noted by now that it's all about being flexible for your reader, and being stubborn about anything you've written isn't going to work.
